---
title: "Exploring and Analysing BRFSS data"
output:
  html_document:
    fig_height: 4
    highlight: pygments
    theme: spacelab
  pdf_document: default
---

## Setup
Load packages ggplot2 and dplyr 
Load data provided 'brfss2013'
### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.RData")
```
* * *

## Part 1: Data
The Behavioral Risk Factor Surveillance System (BRFSS) is the nation's premier system of health-related telephone surveys that collect state data about U.S. residents regarding their health-related risk behaviors, chronic health conditions, and use of preventive services. 
The interviewers collect data from randomly selected adult in a household in conducting BRFSS landline telephone survey and from an adult who participates by using a cellular telephone and resides in a private residence or college housing.
BRFSS data was collectd through the random phone call surveys for non-institutionalized population., therefore the findings are generalized to US population that has either a landline telephone or cellular phone. This study is not generalized to those instituitionalized or those without telephones. 
Since all the data was collected (observed), this study is not based on any random experiment, hence there is no causal inference.

* * *
```{r check the structure of data frame}
str(brfss2013)
```
```{r check dimensions of data frame}
dim(brfss2013)
```
```{r check names of variables}
names(brfss2013)
```

## Part 2: Research questions

**Research quesion 1:**
What is the relationship between education and income level? 

**Research quesion 2:**
What is the impact of number of hours slept on general health of males and females?

**Research quesion 3:**
In year 2013, what percent of males and females got the flushots in the month of November?

* * *

## Part 3: Exploratory data analysis


**Research quesion 1:**

```{r What is the relationship between education and income level? Selected the desired variables.}
  
study1<-brfss2013 %>% 
  select(income2, educa)
```

```{r Filter the observations which have value NA}

study1_NAremoved <- study1 %>% 
  filter(!is.na(educa)) %>% 
  filter(!is.na(income2)) 
```

```{r created a new variable college_graduate to divide the data into two groups - College gratuate or Not college graduate}

study1_NAremoved <- study1_NAremoved %>%
  mutate(College_graduate = ifelse(educa == "College 4 years or more (College graduate)", "Colledge graduate", "Not colledge graduate")) 
```

```{r plot the variables to find the association between education and income }

ggplot(study1_NAremoved, aes(income2, fill = College_graduate))+geom_bar(position = "stack")
```
From above plot it can be configured that income is dependent on the level of education. Those who are College graduate they fall in the higher income range.


* * *


**Research quesion 2:**

```{r What is the impact of number of hours slept on general health of males and females? selected only desired variables.}

study2<-brfss2013 %>% 
  select(sleptim1, physhlth, sex)
```

```{r Filter those observations which have value NA}

study2_NAremoved <- study2 %>%
  filter(!is.na(sleptim1)) %>%
  filter(!is.na(physhlth)) %>%
   filter(!is.na(sex))
```

```{r created a new variable sleep-status to divide the data into two groups - optimum and non optimum. When number of hours of sleep are greater than or equal to 7 and less than 10, otherwise  non optimum. }

study2_NAremoved <- study2_NAremoved %>%
  mutate(sleep_status = ifelse(sleptim1 >= 7 & sleptim1 <= 10, "optimum", "non optimum"))
```

```{r summarized the data by grouping of sleep status and sex}

study2_NAremoved %>% 
  group_by(sleep_status, sex) %>%  summarise(mean_physhlth = mean(physhlth), sd_physhlth = sd(physhlth), n = n())
```

```{r Visualize the impact of sleep status on general health of both the genders male and female.}

ggplot(study2_NAremoved, aes(x = sleep_status, y = physhlth)) + geom_boxplot() + facet_wrap(~sex)


```
The above plots indicate that those who have optimum hours of sleep more than or equal  to 7 hours and less than 10 hours, their physical health is very good for both males and females.


* * *


**Research quesion 3:**

```{r In year 2013, what percent of males and females got the flushots in the month of November? Select desired variable for this study #3.}

study3 <- brfss2013 %>%
  select(imonth, sex, flushot6)
```

```{r Filter observations which have value NA}

study3_NAremoved <- study3 %>%
  filter(!is.na(imonth)) %>%
  filter(!is.na(sex)) %>%
   filter(!is.na(flushot6))
```

```{r filter the observations by month }

study3_November <- study3_NAremoved %>%
  filter(imonth == "November")

```

```{r count the numbers of males and females according to flushot status}

study3_November %>%
    count(flushot6, sex)

```

```{r calculate the percentage of males who got flushots in November 2013}

6554 / (6554 + 9227) 
```
Approx. 41% males got the flushots in November 2013.

```{r calculate the percentage of females who got flushots in November 2013}

10324 / (10324 + 12009)
```
Approx. 46% females got the flushots in November 2013.

```{r plot the observations for the month of November for both genders}

ggplot(study3_November, aes(sex, fill = flushot6)) + geom_bar()
```
This plot and the calculation indicates that indicates that in November 2013, more females got the flushots that males.
